Demo: PhoneLets -- offloading the phone off your phone for energy, cost and network load optimization

### 2014-07-01 paper.tex
submitted version

%make title bold and 14 pt font (Latex default is non-bold, 16 pt)
\title{PhoneLets: offloading the phone off your phone for energy, cost and network load optimization}
\ifnum\anon=1
\author{[Paper: \#XXX]}% \hspace{0.2cm} \ampmtime ]}
\else
\numberofauthors{2}
\author{
\alignauthor Andrius Aucinas\\
\affaddr{University of Cambridge} 
\and
\alignauthor Jon Crowcroft\\
\affaddr{University of Cambridge}
}
\fi
%for single author (just remove % characters)

@@

A modern smartphone has more in common with a computer than a phone, and connectivity to the Internet and surrounding devices (the \emph{Internet of Things}) is tipping the balance even further. Importance of cellular networks, however, remains as they provide almost ubiquitous Internet connectivity around the world. Cellular connectivity has its own problems. This demo shows how some of them can be mitigated by offloading cellular connectivity to a \emph{PhoneLet} - sharing one SIM (Subscriber Identity Module) between nearby devices.

Firstly, it is relatively expensive for the users~\cite{Anonymous:2013ut} and typically the end user needs to obtain a separate subscription for each of his devices connected to the cellular network, e.g. smartphone and tablet as well as broadband connectivity at home and at work. Each device often has its own subscription and fees and sharing a single SIM across them solves the issue.

Secondly, the price is largely driven by the high infrastructure costs which increase with the growing number of network users - the increasing number of connected, periodically communicating clients also increases cellular network load, in the extreme cases causing \emph{signaling storms}~\cite{Yang:wh} that can bring down an entire network. Our approach reduces the number of simultaneously connected clients by sharing the same cellular link across all clients.

Finally, energy consumption of always connected clients is high compared to short-range radio technologies. Sharing a cellular link also helps reduce it for mobile devices. In previous work we analyzed the power consumption of cellular communication~\cite{Aucinas:2013uk} and showed that careful optimization is necessary to prolong the battery life of mobile devices. We discuss in Section~\ref{sec:energy} how PhoneLets can help with such optimization.

\emph{PhoneLets} enable phone offloading by exploiting remote SIM (RSAP) functionality of existing hardware. A cellular module exchange commands with a SIM/UICC module over a wireless communication link rather than direct, physical connection. Some uses of such technology in the past have included automotive as well as electronic currency~\cite{subramanian2003sim}. \emph{PhoneLets} infrastructure establishes and manages secure communication channels between participating entities and allows a device to connect to a cellular network without a physical SIM module being present.

@@

Our design extends the standardized RSAP version by using a SIM server which is not capable of mobile connectivity itself and by using arbitrary communication links to access a remote SIM card (\emph{no SIM} in the mobile client). Our work for this demo focused primarily on the client and SIM server design, and we are using off-the-shelf hardware as the mobile server to demonstrate backwards-compatibility.

\begin{figure}[t!]
\centering
\includegraphics[width=0.9\columnwidth]{figs/sequence}
\vspace{-2mm}
\caption{Communication sequence between client, server and remote SIM card}
\label{fig:sequence}
\vspace{-4mm}
\end{figure}

Mobile Client can obtain network credentials from any server -- a user's mobile phone, a co-located or a remote SIM server -- as long as it can reach it. The key observation is that there are few operations that require access to the SIM card due to its interface speed (it would be infeasible to use it on the data path) and instead it is used to generate encryption keys. The generating keys, however, can never be accessed directly. Part of the network's security however comes from the network authenticating the subscriber for any major operation (e.g. placing or receiving calls, establishing PDP context). The cellular module therefore must always be able to access the remote card using standard Application Protocol Data Unit (APDU) messages (Fig.~\ref{fig:sequence}).

@@

\begin{figure}[t!]
\centering
\includegraphics[width=0.9\columnwidth]{figs/client}
\vspace{-2mm}
\caption{Client prototype}
\label{fig:client}
\vspace{-3mm}
\end{figure}

@@

Cost and network resource savings of using one active subscriber instead of multiple are straightforward, but reduced mobile client energy consumption is an important motivation for \emph{PhoneLets}. We analyze energy consumption of the client in a few example scenarios using our early prototype. The \emph{PhoneLet} itself is not yet energy-optimized, i.e. the controlling RaspberryPI is not meant for mobile uses, we focused on energy comparison of the mobile SIM server - a phone carried by the user from which we offload phone functionality.

\begin{table}[t]
{
\small
\begin{tabular}{| l | c |}
\hline
  \textbf{Use case}         & \textbf{Energy consumption (mA)}  \\ \hline
  Idle - airplane mode      & 5     \\ \hline
  Ifle - BlueTooth          & 6     \\ \hline
  Idle - WiFi               & 8     \\ \hline
  Idle - 3G                 & 10    \\ \hline
  Apps background - WiFi    & 27    \\ \hline
  BlueTooth SIM access      & 28    \\ \hline
  Apps background - 3G      & 40    \\ \hline
  WiFi tethering (idle)     & 80    \\ \hline
\end{tabular}
}
\vspace{-2mm}
\caption{Energy consumption of an idle device with different radios enabled and applications running in background (\emph{Apps background} - Google Apps and Facebook for the test)}
\label{tab:energy}
\vspace{-3mm}
\end{table}

@@

The default BlueTooth implementation of RSAP is not efficient enough - maintaining access alone consumes almost three times the amount of energy of an idle device connected to a 3G network. Instead, we have implemented SIM sharing over WiFi using our custom client and verified that the BlueTooth version is unnecessarily wasteful - once the PhoneLet has registered to the network, there is no communication with the SIM server for as long as the device is idle. For example, a phonecall is successfully received after 30 minutes of no activity. Current Android platform does not allow RSAP to run over WiFi, but based on our measurements we expect energy consumption of an optimized implementation to be very close to that of an idle WiFi channel - 3.5x less than BlueTooth.

Finally, since RSAP can be seen as an alternative to WiFi tethering when the mobile client provides connectivity to other devices, we measured energy consumption of WiFi tethering. With just one idle client it is almost 3x higher than the BlueTooth case. RSAP over WiFi effectively reverses the roles of client and server (\emph{PhoneLet} becomes the server) and similarly as above we expect energy consumption of an optimized WiFi mobile client to be close to 10x less than WiFi tethering.

>>>>

%make title bold and 14 pt font (Latex default is non-bold, 16 pt)
\title{Demo: PhoneLets -- offloading the phone off your phone for energy, cost and network load optimization}
\ifnum\anon=1
\author{[Paper: \#XXX]}% \hspace{0.2cm} \ampmtime ]}
\else
\numberofauthors{2}
\author{
\alignauthor Andrius Aucinas\\
\affaddr{University of Cambridge} 
\and
\alignauthor Jon Crowcroft\\
\affaddr{University of Cambridge}
}
\fi
%for single author (just remove % characters)

@@

A modern smartphone has more in common with a computer than a phone, and connectivity to the Internet and surrounding devices (the \emph{Internet of Things}) is tipping the balance even further. Importance of cellular networks, however, remains as they provide almost ubiquitous Internet connectivity around the world. Cellular connectivity has its own problems. This demo shows how PhoneLets allow one SIM (Subscriber Identity Module) to be shared between nearby devices. This addresses the following problems with cellular connectivity:

Firstly, connectivity is relatively expensive for users~\cite{Anonymous:2013ut}, typically requiring a separate subscription for each device connected to the cellular network, e.g. smartphone and tablet as well as broadband connectivity at home and at work. Using PhoneLets to share a single SIM across devices solves this issue.

Secondly, the increasing number of connected, periodically communicating clients increases infrastructure costs and also increases cellular network load, in extreme cases causing \emph{signaling storms}~\cite{Yang:wh} that can bring down an entire network. PhoneLets reduces the number of simultaneously connected clients by sharing the same cellular link across multiple clients.

Finally, energy consumption of always connected clients is high compared to short-range radio technologies. Sharing a cellular link also helps reduce it for mobile devices.

\emph{PhoneLets} enable phone offloading by exploiting remote SIM functionality of existing hardware. A cellular module exchange commands with a SIM/UICC module over a wireless communication link rather than direct, physical connection. Uses of such technology have included automotive as well as electronic currency~\cite{subramanian2003sim}. \emph{PhoneLets} infrastructure establishes and manages secure communication channels between participating entities and allows a device to connect to a cellular network without a physical SIM module being present.

@@

Our design extends the standardized RSAP (Remote SIM Access Protocol) version by using a SIM server which is not capable of mobile connectivity itself and by using arbitrary communication links to access a remote SIM card (\emph{no SIM} in the mobile client). Our work for this demo focused primarily on the client and SIM server design, and we are using off-the-shelf hardware as the mobile server to demonstrate backwards-compatibility.

\begin{figure}[t!]
\centering
\includegraphics[width=0.9\columnwidth]{figs/sequence}
% \vspace{-2mm}
\caption{Communication sequence between client, server and remote SIM card}
\label{fig:sequence}
% \vspace{-4mm}
\end{figure}

The mobile Client can obtain network credentials from any server -- a user's mobile phone, a co-located or a remote SIM server -- as long as it can reach it. The key observation is that there are few operations that require access to the SIM card due to its interface speed (it would be infeasible to use it on the data path) and instead it is used to generate encryption keys. The generating keys, however, can never be accessed directly. Part of the network's security however comes from the network authenticating the subscriber for any major operation (e.g. placing or receiving calls, establishing PDP context). The cellular module therefore must always be able to access the remote card using standard Application Protocol Data Unit (APDU) messages (Fig.~\ref{fig:sequence}).

@@

\begin{figure}[t!]
\centering
\includegraphics[width=0.9\columnwidth]{figs/client}
% \vspace{-2mm}
\caption{Client prototype}
\label{fig:client}
% \vspace{-3mm}
\end{figure}

@@

Reduced mobile client energy consumption is an important motivation for \emph{PhoneLets}. We analyze energy consumption of the client in a few example scenarios using our early prototype. The \emph{PhoneLet} itself is not yet energy-optimized, i.e. the controlling RaspberryPI is not meant for mobile uses, so we focused on energy comparison of the mobile SIM server - a phone carried by the user from which we offload phone functionality.

\begin{table}[t]
{
\small
\begin{tabular}{| l | c |}
\hline
  \textbf{Use case}         & \textbf{Energy consumption (mA)}  \\ \hline
  Idle - airplane mode      & 5     \\ \hline
  Ifle - BlueTooth          & 6     \\ \hline
  Idle - WiFi               & 8     \\ \hline
  Idle - 3G                 & 10    \\ \hline
  Apps background - WiFi    & 27    \\ \hline
  BlueTooth SIM access      & 28    \\ \hline
  Apps background - 3G      & 40    \\ \hline
  WiFi tethering (idle)     & 80    \\ \hline
\end{tabular}
}
% \vspace{-2mm}
\caption{Energy consumption of an idle device with different radios enabled. (Apps background = Google apps and Facebook)}
\label{tab:energy}
% \vspace{-3mm}
\end{table}

@@

The default BlueTooth implementation of RSAP is not efficient enough - maintaining access alone consumes almost three times the amount of energy of an idle device connected to a 3G network. Instead, we have implemented SIM sharing over WiFi using our custom client and verified that the BlueTooth version is unnecessarily wasteful - once the PhoneLet has registered to the network, there is no communication with the SIM server for as long as the device is idle. For example, a phonecall is successfully received after 30 minutes of no activity. The current Android platform does not allow RSAP to run over WiFi, but based on our measurements we expect energy consumption of an optimized implementation to be very close to that of an idle WiFi channel - 3.5x less than BlueTooth.

Finally, we measured energy consumption of WiFi tethering. With just one idle client it is almost 3x higher than the BlueTooth case. RSAP is effectively an alternative to WiFi tethering when the mobile client provides connectivity to other devices. So similarly to above, we expect energy consumption of an optimized WiFi mobile client to be close to 10x less than WiFi tethering.
